How has the Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985

### 2023-02-14 Submission/Re-submission/draft/draft_resubmission_shorten.rmd
with changes

---
title: How has the Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both co-authors contributed equally to the article. Names are ordered alphabetically. Livio is a PhD Candidate in Sociology and affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences and affiliated to the Centre for International Environmental Studies.]
date: 13 February 2023
output:
  pdf_document: default
  latex_engine: xelatex
  fig_caption: yes
  html_document:
    df_print: paged
fontsize: 12pt
mainfont: Times New Roman
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage[T1]{fontenc}
   - \usepackage[utf8]{inputenc}
   - |
    \providecommand{\keywords}[1]{\textbf{Keywords:} #1}
    \providecommand{\wordcount}[1]{\textbf{Word Count:}#1}
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: The Amazon is often approached as a transnational environmental problem without a discussion of how key actors see it differently in space and time. We combine theories of problem-representation and transnationalism to analyze where, when, and how the Brazilian Amazon has been constructed as a problem. We develop a framework that explains variation in transnational problem-constructions based on speech setting and problem pressingness. To test our framework, we use supervised machine learning to classify statements about the Amazon in 6240 presidential speeches since 1985. We find that when presidents speak outside (or inside) the Amazon and environmental problems are less (or more) pressing, presidents are likelier to highlight issues of environmental conservation (or economic integration). Furthermore, we show how we are moving from a period when the binary between seeing the Amazon as an environmental or an economic issue has been overcome, to a moment when diverse problem-constructions are deployed.

@@

Scholars studying the Brazilian Amazon often identify different governments throughout Brazilian history as proponents of a single view of the region, its people, and forest [see @drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. While the military dictatorship (1964-85) and Bolsonaro’s presidency (2019-2022) are, for example, associated with understanding the Amazon as issues of national sovereignty and economic integration, Sarney’s (1985-1990) and Lula’s (2003-2010) presidencies are associated with environmental conservation. Yet, such representations advance the view that governments understood and acted consistently towards the Amazon as a specific problem, masking government diversity of discursive constructions and policies. In this article, we show where, when, and how the Brazilian Amazon has been constructed as a problem in presidential discourses since 1985.

By combining @bacchi1999’s theory of problem representation and @keck1998's theory of transnationalism, we place the way the objects of policies are constructed in discourse at the center of our analysis. We offer a framework that explains variation in discursive problem-constructions depending on speech setting and problem pressingness as actors participate and shape transnational politics. We focus on transnational speeches, rather than policies, because they allow us to capture the diversity of ways the Amazon, as a policy object that demands dedicated attention, has been discursively constructed as a problem. While many analyses of Brazilian environmental discourse and policies intend to dissect its national and international nature, they rarely adopt methodological strategies that distinguish how important actors behave differently in both scenarios [@hochstetler2007]. Our framework and methodology help identify and account for these variations. We show, for instance, that when speaking to farmers in Cuiaba, one of Brazil’s agricultural powerhouses, in a moment when the environmental problem is pressing, it is likely that any president would sidestep environmental issues, as President Lula did in the opening epigraph.

To answer our question, we create a dataset containing 6240 official speeches by all Brazilian presidents since 1985. Next, we identify all instances where the presidents mention the Amazon. Using an inductive approach, we develop a codebook to categorize how each statement constructs the Amazon as one or multiple problems related to national sovereignty, economic integration, social development, or environmental conservation. Then, we manually code a random subset of the statements and use supervised machine learning to label the remaining. Finally, we conduct a descriptive and inferential analysis of the coded data.

We have four main findings. First, we find that presidents adopt different Amazonian constructions  in different settings according to how they expect to leverage audiences for symbolic and/or  material gains. Second, presidents often focus on alternative problem-constructions when general aligned problems become more pressing; whereas they boast about positive outcomes when aligned problems are less pressing. Third, we also find that high-profile environmental related events, which receive strong international media attention, drive up the importance of the Amazon as a general topic in presidential speeches. Alternatively, relevant Amazonian-related policies  appear to not to attract much attention. Finally, we find that while all presidents deploy multiple problem-constructions across space and over time, we are now in a period when the dictum between understanding the Amazon as an environment or as an economic issue has been overcome, at least discursively. Conceptually, our research shows how focusing on problem-constructions highlight what certain objects of policy are assumed to be, and when they are ignored or privileged, based on the transnational setting of the speech and the pressingness of aligned problems.

@@

We argue that the reason behind these monolithic representations of governments in literature about the Amazon is conceptual: many studies of the region contain an explicit or implicit assumption of what the “Amazon problem” is. There is more variation within a specific government, though, than suggested by the policy cycle approach. Social scientists have shown how governments have contradictory policies [@lopez2023; @fairfield2017; @poncedeleon2021]. In the case of the Amazon, it has often been seen as a key biome in global climate-mitigation, and consequently, as an environmental problem. If the Amazon is thought of as an “environmental” problem, then it demands an “environmental” solution. However, in order to build a more comprehensive account of the Amazon as forest, region, and peoples we need a framework that assumes the possibility of varied problem-constructions beyond Amazonian environmental nature.

@@

Following Bacchi, we place discourse about policy objects in the center of our conceptual framework. We define policy objects as political objects that demand dedicated political attention. We argue that by deciding to speak about a policy object, governments are making a deliberate choice. When speaking about policy objects, governments make a choice to represent, or construct it, as a specific problem. The problem might be either explicitly stated, or implicitly hidden in the underlined solution. This is what we label as problem-constructions.

Presidential discourses are a good instance to identify problem-constructions over time. While the relationship between presidential and policy may not be linearly causal, speeches have the power to introduce, define, and justify public policy, as well as shape issue perception to broad audiences [@zarefsky2004]. This is particularly true in the Brazilian context where the executive branch of government, led by the president, has historically been the primary player in shaping public policy, especially after the 1988 constitution[@macaulay2017]. In  Brazil’s hybrid system, often referred to Coalitional Presidencialism [@couto2021], presidents build legislative support for a broad agendas through the distribution of positions in the state bureaucracies. This means that presidents have enough power and support to implement the agenda they deem appropriate. We thus consider presidential speeches an adequate instance to identify various problem-constructions for the same policy object.

@@

@putnam1988 argues that the outcomes of international negotiations lie within the overlap between the interests (and pressure) of domestic and international groups. At the national level, interest groups pressure the government to adopt favorable policies. Internationally, the government seeks to fulfill domestic preferences “while minimizing the adverse consequences of foreign developments” [@putnam1988 p. 434]. While the theory has been expanded to cover multi-levels games [see @collinson1999], its core remains the same: the outcome of negotiations should lie within the overlap of diverse interest groups. Consequently, presidential speeches are not necessarily contradictory across venues but a common denominator of the interest of different groups. Adopting two-level games to explain American presidential speeches, @calderwood2019 and @calderwood2020 find supporting evidence of this theory.

@@

While their theory is developed in the context of value-based advocacy groups, this is non-state actors, the overall view of international politics as structured and structuring, should hold for state actors as well. Important participants of transnational networks, as presidents, have a myriad of tactics to introduce, define, persuade, socialize, and pressure when interacting with other actors. They can create politically meaningful information and/or move it where they believe it can have an impact [@keck1998, p. 24]. Or, they can leverage institutions to realize symbolic or material gains. In this sense, the positions of actors as components of a transnational network derive from complex interactions in which their identities and goals are malleable and can vary depending on the space and time of the interaction.

@@

Presidents construct problems and attempt to inform and/or leverage transnational audiences for symbolic or material gains according to their expectations of that audience. We conceptualize spaces in terms of settings. Settings are levels of aggregation for transnational audiences that make sense in light of the object of policy under study. This indicates that how presidents construct policy objects as problems can vary, for instance, within the country — speaking inside the Amazon region versus outside — because presidents identify these audiences as sources of different types of support and demand. The same holds when speaking outside the country — in Amazonian countries, non-Amazonian countries, and international summits — where expectations and identities of audiences could diverge completely from that of domestic actors. In each one of these settings different problems might be more pressing and the president's expectations about what can be leveraged for material and symbolic gains might drive how a problem is implicitly or explicitly constructed.

@@

How presidents construct policy objects can also vary in time because the pressingness of certain problems changes as new information arises. We conceptualize time as the perceived pressingness of a broad and related issue to those surrounding a policy object. This indicates, for instance, that as climate change becomes more pressing and widespread as a broad issue, it is more likely that actors will consider climate change when discussing various forms of public policy. It is likely that when speaking about subsidies for fossil fuel companies in the 1980s, presidents would not mention the dangers of climate change. Conversely, today, if fossil fuel subsidies are discussed, climate change will likely be an issue given the increased pressingness of the topic in comparison to the 1980s.

@@

Our conceptual framework, thus, combines Bacchi’s theory of problem-representation and Keck and Sikkink’s theory of transnationalism to account for diverse problem-constructions. It helps to advance the literature about the Amazon in two ways: We provide an outlook of the region, its people and forest beyond its environmental nature by focusing on different problem-constructions. In doing so, we avoid assuming what type of problem the Amazon is and, instead, focus on how it has been represented as a problem. Furthermore, many analyses of Brazilian environmental policies dissect its national and international nature but fail to adopt methodological strategies that distinguish how important actors behave differently in both scenarios [@hochstetler2007]. Our focus on presidential speeches affords us analytical leverage to identify transnational variation, as speeches take place in multiple settings over time. Finally, presidential speeches in Brazil have been studied for topics such as race relations [@silva2019] and inequality [@grangeia2017]. When it comes to the environment, studies tend to focus primarily  on Bolsonaro’s discourse [@marquardt2022; @mendes2022] or policies [see @capelari2023]. We argue, however, that a longitudinal view across Brazilian presidents is important to understand how much continuity and change there really is.

@@

We use a codebook developed through an inductive process and informed by the literature to categorize Amazonian statements into four problem-constructions: national sovereignty, economic integration, environmental conservation, and social development. While the three former are in the literature, we argue a fourth one, social development, exists.

*National Sovereignty* 
In the process of securing Amazonian borders since the 18th century, Brazil thwarted “the imperial ambitions of France, Britain, the United States, Belgium, Bolivia, and Peru” [@hecht1990, p. 8]. While the Brazilian military diplomacy was very successful, the process did not come without its traumas as, for example, the negotiations with Bolivia in 1902 to secure the state of Acre [@hecht1990]. These traumas were referred to, and offered context, to Amazonian discourses and policies that highlighted Brazil’s sovereignty over the Amazon [@rajao2020]. More recently, as we move to a world where non-state actors gain importance in international politics [@andonova2014; @keck1998], sovereignty-related problems become more varied. Threats to national sovereignty, consequently, broaden from nation-states to a wide set of actors. The sovereignty problem-construction advances the view that the Brazilian Amazon belongs to Brazil and any foreign or non-state presence in the region can be part of a broad strategy to take the region. The policy solutions to the issue of sovereignty included the monitoring of the borders and strict regimes related to entry into the region.

@@

*Environmental Conservation*
The rapid economic changes in the region from the 1960s to the 1980s were matched with the birth of environmental institutions [@drummond2006]. The creation of these institutions stemmed from the impression of the lack of control over the environment following the efforts to economically integrate in the region [@acker2021; @capobianco2021; @hecht1990]. This process accelerated in the late 1980s with the birth of modern environmentalism [@viola1987]. The environmental conservation problem-construction emphasizes that the Amazon should be preserved by the creation of protected areas and deforestation should be halted. The policy solution implies, for example, further investments in command-and-control institutions, the valuation of standing ecosystems through incentive schemes, and the creation of protected areas.

@@

To operationalize settings, we extract the location of the speech and create four distinct settings: Amazonian states within Brazil, non-Amazonian states within Brazil, Amazonian countries, and non-Amazonian countries. Each one of these settings encompasses audiences that can have different expectations in reference to our policy object, the Brazilian Amazon. When speaking at Amazonian states presidents face audiences that likely have distinct social, economic, or environmental demands in relation to the Brazilian Amazon. Such expectations might be distinct from the ones audiences in Brazil, but outside of the Amazonian region, have for the same policy object. During visits to Amazonian countries, presidents may view them as peers facing similar challenges and adjust their speeches accordingly. Conversely, when speaking in non-Amazonian countries or at international events such as the UNFCCC or visits to the US or France, presidents may tailor their speeches to align with what they perceive as possible symbolic or material gains from the audience.

@@

We also add two control variables in the models: election years and peak years. In the case of election years, we control for the possibility of presidents constructing the Amazon in systematically different ways in those years. In the case of peak years, these are years in which the Amazon was disproportionately more mentioned as a share of all presidential speeches. We calculate these by capturing years in which mentions of the Amazon are above one standard deviation from their mean share.

@@

The following models (2-5) address the question of how the Amazon is constructed as a problem in different settings. In terms of environmental conservation (model 2), we observe that when speaking at non-Amazonian countries, be it in international summits, bilateral meetings, or fora, presidents are much more likely to construct the Amazon as an issue of environmental conservation, in relation to when they speak within Amazonian states in Brazil (reference category). To a lesser extent, this also holds for when presidents speak in non-Amazonian states within Brazil where they are more likely to construct the Amazon as an issue of environmental conservation. While the results for economic integration (model 3) and social development (model 4) problem-constructions are not as robust, they are strong enough to suggest the opposite direction: both are less likely outside Brazil. Finally, when it comes to national sovereignty (model 5), the problem-constructions are more likely to take place when presidents speak at non-Amazonian states in Brazil than when they are speaking elsewhere.

@@

## 4.2 Amazonian speeches in time

Looking at speeches that mention the Amazon in time (the dependent variable of model 1) reviews several interesting trends. Figure 2, below, shows the proportion of speeches that mentions the Amazon in relation to all speeches in a year. By investigating its development in time, we can identify instances when the Amazon in itself becomes more pressing as a problem, and when not. On average, presidents mention the Amazon in 15.5% of their speeches ^[For comparison purposes, the averages for other policy objects in the same corpus of speeches are inequality (13.7%), criminality (17.3%), inflation (19.1%), and unemployment (13.3%).]. This suggests the Amazon is a topic of relative high importance for presidents generally. However, we observe four peak years (one standard deviation from the mean) in the figure: 1989, 1992, 2009, and 2019. We find that these peak years coincide with high-profile environmental events, but not with domestic policy adoption.

@@

In 1989 the Amazon appeared in 32% of all speeches. This coincides with the brutal murder of the environmental activist Chico Mendes in the last days of 1988. The incident caught unprecedented transnational media attention and then President Sarney (1985-1989) responded to this with a set of policies to address deforestation [@capobianco2021]. One of the responses was the offer and acceptance to host the 1992 Earth Summit [@keck1998], which also led to a peak in 1992. During the summit, we saw the realization of the first transnational partnership for the Amazon, the Pilot Programme of the G7 for the Protection of Rainforest (PPG7), which brought a high number of financial resources to the region for public policy implementation [@capobianco2021].

Throughout the rest of the 1990s and 2000s, during the Cardoso (1995-2002) and Lula presidencies, the Amazon as a topic did not diverge much from the average. This is aligned with our model as the pressingness of the environmental problem was high during the period. In 2009, however, we observed another peak which coincided with the Copenhagen Summit. Lula led the delegation to the summit with a strong climate leadership self-image of “we deliver” [@franchini2019], in a moment when deforestation rates have been decreasing since 2005. From 2010 to 2016, except for 2015 when the Paris summit took place, we see a general decrease in mentions of the Amazon in all official presidential speeches. In this period,  the pressingness of environmental problems was increasing as deforestation rates were rising. The period was also marked by political and economic instability which culminated in the impeachment of President Rousseff in 2016. By 2016 mentions of Amazon in official speeches went down to 6%, their lowest share since 1985.

@@

The figure 2, above, also shows that the frequency in which the Amazon is mentioned, as a general topic in speeches, is not related by proposed policy solutions. In 2000, for instance, Brazil adopted a strong legislation regulating the creation of protected areas [@silva-muller2022]. In 2004, President Lula decreed the creation of Brazil’s Plan to Combat and Control Deforestation in the Brazilian Amazon. In 2008, Brazil announced the Amazon Fund, the largest fund of this kind ever created. Finally, in 2012, the revision of the Forest Code, debated in 2010 and 2011, pardoned thousands of illegal deforesters in the Amazon [@sant2021]. Yet, these policy landmarks do not seem to be systematically meaningful for when Amazon is mentioned as a general problem more often.

@@

Pure economic integration constructions were the most frequent from 1985 to 2009. This is especially pertinent during Cardoso’s presidency (1995-2002), when construction of the Amazon as an issue of economic integration increased while environmental conservation constructions decreased. This was a period when the pressingness of the environmental problem grew as deforestation rates were increasing, so presidents were more likely to avoid environmental conservation constructions. When Lula took office (2003-2010), construction of the Amazon as an issue of environmental conservation and social development increased. During this period, the pressingness of environmental problems decreased along with deforestation rates, so presidents are more likely to employ environmental conservation constructions. From around 2010 until the mid-2010s, especially during the Rousseff (2011-2016) administration, we see a reversal at the top whereas environmental conservation and social development problem-constructions surpass economic integration. However, this reversal was not long-standing. With the presidencies of Temer (2016-2018) and Bolsonaro (2019-2022), economic integration constructions become the most favored again.

>>>>

---
title: How has the Amazon been constructed as a problem? Presidential speeches and transnational politics since 1985
author:
- Livio Silva-Muller
- Henrique Sposito^[Both authors contributed equally to the article. Names are ordered alphabetically. Livio is a PhD Candidate in Sociology and affiliated to the Albert Hirschman Centre on Democracy. Henrique is a PhD Candidate in International Relations and Political Sciences and affiliated to the Centre for International Environmental Studies.]
date: 13 February 2023
output:
  pdf_document: default
  latex_engine: xelatex
  fig_caption: yes
  html_document:
    df_print: paged
fontsize: 12pt
mainfont: Times New Roman
header-includes:
   - \usepackage{floatrow}
   - \floatsetup[figure]{capposition=top}
   - \usepackage{float}
   - \floatplacement{figure}{H}
   - \usepackage[T1]{fontenc}
   - \usepackage[utf8]{inputenc}
   - |
    \providecommand{\keywords}[1]{\textbf{Keywords:} #1}
    \providecommand{\wordcount}[1]{\textbf{Word Count:}#1}
always_allow_html: TRUE
bibliography: references.bib
csl: chicago.csl
abstract: The Amazon is often approached as a transnational environmental problem without a discussion of how key actors see it differently in space and time. We combine theories of problem-representation and transnationalism to analyze where, when, and how the Brazilian Amazon has been constructed as a problem. We develop a framework that explains variation in transnational problem-constructions based on speech setting and problem pressingness. To test our framework, we use supervised machine learning to classify statements about the Amazon in 6240 presidential speeches since 1985. We find that when presidents speak outside (or inside) the Amazon and environmental problems are less (or more) pressing, presidents are likelier to highlight issues of environmental conservation (or economic integration). Furthermore, we show how we are moving from a period when the Amazon as a binary of either environmental or an economic issue has been overcome, to a moment when diverse problem-constructions are deployed.

@@

Scholars studying the Brazilian Amazon often identify different governments throughout Brazilian history as proponents of a single view of the region, its people, and forest [see @drummond2006; @padua2012; @franchini2019; @capobianco2019; @pereira2021]. While the military dictatorship (1964-85) and Bolsonaro’s presidency (2019-2022) are, for example, associated with understanding the Amazon as issues of national sovereignty and economic integration, Sarney’s (1985-1990) and Lula’s (2003-2010) presidencies are associated with environmental conservation. Yet, such representations advance the view that governments act consistently towards the Amazon as a specific problem, masking government diversity of in policies and discourse. In this article, we show where, when, and how the Brazilian Amazon has been constructed as a problem in presidential discourses since 1985.

By combining @bacchi1999’s theory of problem representation and @keck1998's theory of transnationalism, we place the way the objects of policies are constructed in discourse at the center of our analysis. We offer a framework that explains variation in discursive problem-constructions depending on speech setting and problem pressingness as actors participate and shape transnational politics. We focus on speeches, rather than policies, because they allow us to capture the diversity of ways the Amazon, as a policy object that demands dedicated attention, has been discursively constructed as a problem. While many analyses of Brazilian environmental discourse and policies intend to dissect its national and international nature, they rarely adopt methodological strategies that distinguish how important actors behave differently in both scenarios [@hochstetler2007]. Our framework and methodology help identify and account for these variations. We show, for instance, that when speaking to farmers in Cuiaba, one of Brazil’s agricultural powerhouses, in a moment when the environmental problem is pressing, it is likely that any president would sidestep environmental issues, as President Lula did in the opening epigraph.

To answer how the Amazon has been constructed as a problem, we create a dataset containing 6240 official speeches by all Brazilian presidents since 1985. Next, we identify all instances where the presidents mention the Amazon. Using an inductive approach, we develop a codebook to categorize how each statement constructs the Amazon as one or multiple problems related to national sovereignty, economic integration, social development, or environmental conservation. Then, we manually code a random subset of the statements and use supervised machine learning to label the remaining. Finally, we conduct a descriptive and inferential analysis of the coded data.

We have four main findings. First, we find that presidents adopt different constructions of the Amazon in different settings according to how they expect to leverage audiences for symbolic and/or  material gains. Second, presidents often focus on alternative problem-constructions when problems become more pressing; whereas they boast about positive outcomes when problems are less pressing. Third, we find that high-profile environmental related events, which receive strong international media attention, increase the importance of the Amazon in presidential speeches. Alternatively, relevant Amazonian-related policies appear to not to attract much attention. Finally, we find that while all presidents deploy multiple problem-constructions across space and over time, we are now in a period when the dictum between understanding the Amazon as an environment or as an economic issue has been overcome, at least discursively. Conceptually, our research shows how focusing on problem-constructions highlight what certain objects of policy are assumed to be, and when they are ignored or privileged, based on the transnational setting of the speech and the pressingness of aligned problems.

@@

We argue that the reason behind these monolithic representations of governments in literature about the Amazon is conceptual: many studies of the region contain an explicit or implicit assumption of what the “Amazon problem” is. There is more variation within a specific government, though, than suggested by the policy cycle approach. Social scientists have shown how governments have contradictory policies [@lopez2023; @fairfield2017; @poncedeleon2021]. The Amazon has often been seen as a key biome in global climate-mitigation, and consequently, as an environmental problem. If the Amazon is thought of as an “environmental” problem, then it demands an “environmental” solution. However, to build a more comprehensive account of the Amazon as forest, region, and peoples we need a framework that assumes the possibility of varied problem-constructions beyond Amazonian environmental nature.

@@

Following Bacchi, we place discourse about policy objects at the center of our conceptual framework. We define policy objects as objects that demand dedicated political attention and argue that by deciding to speak about a policy object, governments are making a deliberate choice to represent, or construct it, as a specific problem. The problem might be either explicitly stated, or implicitly hidden in the underlined solution. This is what we label as problem-constructions.

Presidential discourses are a good instance to identify problem-constructions over time. While the relationship between presidential speeches and policy may not be linearly causal, speeches have the power to introduce, define, and justify public policy, as well as shape issue perception to broad audiences [@zarefsky2004]. This is particularly true in the Brazilian context where the executive branch of government, led by the president, has historically been the primary player in shaping public policy, especially after the 1988 constitution[@macaulay2017]. In Brazil’s hybrid system, often referred to Coalitional Presidencialism [@couto2021], presidents build legislative support for a broad agenda through the distribution of positions in the state bureaucracies. This means that presidents have enough power and support to implement the agenda they deem appropriate. We thus consider presidential speeches an adequate instance to identify various problem-constructions for the same policy object.

@@

@putnam1988 argues that the outcomes of international negotiations lie within the overlap between the interests (and pressure) of domestic and international groups. At the national level, interest groups pressure the government to adopt favorable policies. Internationally, the government seeks to fulfill domestic preferences “while minimizing the adverse consequences of foreign developments” [@putnam1988 p. 434]. While the theory has been expanded to cover multi-levels games [see @collinson1999], its core remains the same: the outcome of negotiations should lie within the overlap of diverse interest groups. Consequently, presidential speeches are not necessarily contradictory across venues but a common denominator of the interests of different groups. Adopting two-level games to explain American presidential speeches, @calderwood2019 and @calderwood2020 find supporting evidence of this theory.

@@

While their theory is developed in the context of value-based advocacy groups, hence non-state actors, the overall view of international politics as structured and structuring, should hold for state actors as well. Important participants of transnational networks, as presidents, can use myriad of tactics to introduce, define, persuade, socialize, and pressure when interacting with other actors. They can create politically meaningful information and/or move it where they believe it can have an impact [@keck1998, p. 24]. They can also leverage institutions to realize symbolic or material gains. In this sense, the positions of actors in transnational networks derive from complex interactions in which their identities and goals are malleable and can vary depending on the space and time of the interaction.

@@

Presidents construct problems and attempt to inform and/or leverage transnational audiences for symbolic or material gains according to their expectations of that audience. We conceptualize spaces in terms of settings. Settings are levels of aggregation for transnational audiences that make sense in light of the object of policy under study. This indicates that how presidents construct policy objects as problems can vary, for instance, within the country — speaking inside the Amazon region versus outside — because presidents identify these audiences as sources of distinct types of support and demand. The same holds true when speaking abroad — in Amazonian countries, non-Amazonian countries, and at international summits — where presidents’ expectations and identities of an audiences could diverge completely from that of domestic actors. In each one of these settings different problems might be more pressing and the president's expectations about what can be leveraged for material and symbolic gains might influence how a problem is implicitly or explicitly constructed.

@@

How presidents construct policy objects can also vary over time because the pressingness of certain problems changes as new information arises. We conceptualize time as the perceived pressingness of a broad and related issue to those surrounding a policy object. This indicates, for instance, that as climate change becomes more pressing and mainstreamed as a broad issue, it is more likely that actors will consider it when discussing various forms of public policy. It is likely that when speaking about subsidies for fossil fuel companies in the 1980s, presidents would not mention the dangers of climate change, whilst today, climate change would likely be mentioned given the increased global attention.

@@

Our conceptual framework, thus, combines Bacchi’s theory of problem-representation and Keck and Sikkink’s theory of transnationalism to account for diverse problem-constructions. We help to advance the literature about the Amazon in two ways. First, we provide an outlook of the region, its people and forest beyond its environmental issues by focusing on different problem-constructions. In doing so, we avoid assuming what type of problem the Amazon is and, instead, focus on how it has been represented as a problem. Second, many analyses of Brazilian environmental policies dissect its national and international nature but do not adopt methodological strategies that distinguish how important actors behave differently in both scenarios [@hochstetler2007]. Our focus on presidential speeches affords us analytical leverage to identify transnational variation, as speeches take place in multiple settings over time. Finally, presidential speeches in Brazil have been studied for topics such as race relations [@silva2019] and inequality [@grangeia2017]. When it comes to the environment, studies tend to focus primarily on Bolsonaro’s discourse [@marquardt2022; @mendes2022] or policies [see @capelari2023]. We argue, however, that a longitudinal view across Brazilian presidents is important to understand how much continuity and change there really is.

@@

We use a codebook developed through an inductive process and informed by the literature to categorize Amazonian statements into four problem-constructions: national sovereignty, economic integration, environmental conservation, and social development. While the three formers are present in the literature, we argue a fourth one exists in speeches: social development.

*National Sovereignty* 
In the process of securing Amazonian borders since the 18th century, Brazil thwarted “the imperial ambitions of France, Britain, the United States, Belgium, Bolivia, and Peru” [@hecht1990, p. 8], and secured the Amazon as Brazilian. While military diplomacy was very successful, the process did not come without its issues and sovereignty in the Amazon became one of the main military concerns[@rajao2020]. More recently, as we move to a world where non-state actors gain importance in international politics [@andonova2014; @keck1998], sovereignty-related problems become more varied. Threats to national sovereignty, consequently, broaden from nation-states to a wide set of actors. The sovereignty problem-construction advances the view that the Brazilian Amazon belongs to Brazil and any foreign or non-state presence in the region can be part of a broad strategy to take the region. The policy solutions to the issue of sovereignty included the monitoring of the borders and strict regimes related to entry into the region.

@@

*Environmental Conservation*
The rapid economic changes in the region from the 1960s to the 1980s were matched with the birth of domestic environmental institutions [@drummond2006]. The creation of these institutions stemmed from the impression of the lack of control over the environment following the efforts to economically integrate in the region [@acker2021; @capobianco2021; @hecht1990]. This process accelerated in the late 1980s with the birth of modern environmentalism [@viola1987]. The environmental conservation problem-construction emphasizes that the Amazon should be preserved by the creation of protected areas and deforestation should be halted. The policy solution implies, for example, further investments in command-and-control institutions, the valuation of standing ecosystems through incentive schemes, and the creation of protected areas.

@@

To operationalize settings, we extract the location of the speech and create four distinct settings: Amazonian states within Brazil, non-Amazonian states within Brazil, Amazonian countries, and non-Amazonian countries. Each one of these settings encompasses audiences that can have different expectations in reference to our policy object, the Brazilian Amazon. When speaking in Amazonian states presidents face audiences that likely have distinct social, economic, or environmental demands in relation to the Brazilian Amazon. Such expectations might be distinct from the ones’ audiences in non-Amazonian Brazilian States. Furthermore, during visits to Amazonian countries, presidents may perceive them as peers facing similar challenges and adjust their speeches accordingly. Conversely, when speaking in non-Amazonian countries or at international events such as the UNFCCC or visits to the US or France, presidents may tailor their speeches to align with what they perceive as possible symbolic or material gains from the audience.

@@

We also add two control variables in the models: election years and peak years. In the case of election years, we control for the possibility of presidents constructing the Amazon in systematically different ways in those years. In the case of peak years, these are years in which the Amazon was mentioned more often as a share of all presidential speeches. We calculate these by capturing years in which mentions of the Amazon are above one standard deviation from their mean share.

@@

The following models (2-5) address the question of how the Amazon is constructed as a problem in different settings. In terms of environmental conservation (model 2), we observe that when speaking in non-Amazonian countries, be it in international summits, bilateral meetings, or fora, presidents are much more likely to construct the Amazon as an issue of environmental conservation, in relation to when they speak within Amazonian states in Brazil (reference category). To a lesser extent, this also holds for when presidents speak in non-Amazonian states within Brazil where they are more likely to construct the Amazon as an issue of environmental conservation. While the results for economic integration (model 3) and social development (model 4) problem-constructions are not as robust, they are strong enough to suggest the opposite direction: both are less likely outside Brazil. Finally, when it comes to national sovereignty (model 5), the problem-constructions are more likely to take place when presidents speak at non-Amazonian states in Brazil than when they are speaking elsewhere.

@@

## 4.2 Amazonian speeches over time

Looking at speeches that mention the Amazon over time (the dependent variable of model 1) reviews several interesting trends. Figure 2, below, shows the proportion of speeches that mentions the Amazon in relation to all speeches in a year. By investigating its development in time, we can identify instances when the Amazon in itself becomes more pressing as a problem, and when not. On average, presidents mention the Amazon in 15.5% of their speeches ^[For comparison purposes, the averages for other policy objects in the same corpus of speeches are inequality (13.7%), criminality (17.3%), inflation (19.1%), and unemployment (13.3%).]. This suggests the Amazon is a topic of relative high importance for presidents generally. However, we observe four peak years (one standard deviation from the mean) in the figure: 1989, 1992, 2009, and 2019. We find that these peak years coincide with high-profile environmental events, but not with domestic policy adoption.

@@

In 1989 the Amazon appeared in 32% of all speeches. This coincides with the brutal murder of the environmental activist Chico Mendes in the last days of 1988. The incident caught unprecedented transnational media attention and then President Sarney (1985-1989) responded to this with a set of policies to address deforestation [@capobianco2021]. One of the responses was to host the 1992 Earth Summit [@keck1998], which also led to a peak in 1992. During the summit, we saw the realization of the first transnational partnership for the Amazon, the Pilot Programme of the G7 for the Protection of Rainforest (PPG7), which brought a high number of financial resources to the region for public policy implementation [@capobianco2021].

Throughout the rest of the 1990s and 2000s, during the Cardoso (1995-2002) and Lula presidencies, there were no peaks. This is aligned with our model as the pressingness of the environmental problem was high during the period. In 2009, however, we observed another peak which coincided with the Copenhagen Summit. Lula led the delegation to the summit with a strong climate leadership self-image of “we deliver” [@franchini2019], in a moment when deforestation rates have been decreasing since 2005. From 2010 to 2016, except for 2015 when the Paris summit took place, we see a general decrease in mentions of the Amazon in all official presidential speeches. In this period,  the pressingness of environmental problems was increasing as deforestation rates were rising. The period was also marked by political and economic instability which culminated in the impeachment of President Rousseff in 2016. By 2016 mentions of Amazon in official speeches went down to 6%, their lowest share since 1985.

@@

The figure 2, above, also shows that the frequency at which the Amazon is mentioned, as a general topic in speeches, is not related to proposed policy solutions. In 2000, for instance, Brazil adopted a strong legislation regulating the creation of protected areas [@silva-muller2022]. In 2004, President Lula decreed the creation of Brazil’s Plan to Combat and Control Deforestation in the Brazilian Amazon. In 2008, Brazil announced the Amazon Fund, the largest fund of this kind ever created. Finally, in 2012, the revision of the Forest Code, debated in 2010 and 2011, pardoned thousands of illegal deforesters in the Amazon [@sant2021]. Yet, these policy landmarks do not seem to be systematically meaningful for when Amazon is mentioned as a general problem more often.

@@

Pure economic integration constructions were the most frequent from 1985 to 2009. This is especially pertinent during Cardoso’s presidency (1995-2002), when construction of the Amazon as an issue of economic integration increased while environmental conservation constructions decreased. This was a period when the pressingness of the environmental problem grew as deforestation rates were increasing, so presidents were more likely to avoid environmental conservation constructions. When Lula took office (2003-2010), construction of the Amazon as an issue of environmental conservation and social development increased. During this period, the pressingness of environmental problems decreased along with deforestation rates, so presidents were more likely to employ environmental conservation constructions. From around 2010 until the mid-2010s, especially during the Rousseff (2011-2016) administration, we see a reversal at the top whereas environmental conservation and social development problem-constructions surpass economic integration. However, with the presidencies of Temer (2016-2018) and Bolsonaro (2019-2022), economic integration constructions become the most favored again.
